noam_beta_diversity

### 2024-08-29 noam_beta_diversity.Rmd

not_algae <- c("ascidian", "bare_space", "coral", "coral_rubble", "corallimorpharia", "heteractis_sp", "millepora_platyphylla", "no_data", "sand", "sacrophyton_sp", "shell_debris", "soft_coral", "sponge", "tridacna_sp", "zooxanthellae", "row_sums")

@@

```{r fringing bray curtis}
fringing_species_matrix <- as.matrix(fringing[ , -which(names(fringing) %in% meta_cols)])
fringing_species_matrix <- fringing_species_matrix/100 # convert from percent to proportion
rownames(fringing_species_matrix) <- paste(fringing$year,"-", fringing$location, sep = "")
fringing_rows_to_remove <- rowSums(fringing_species_matrix) == 0 # remove rows with no algal species observed - thought this was done already
fringing_species_matrix <- fringing_species_matrix[!fringing_rows_to_remove,] 
fringing_species_matrix <- asin(sqrt(fringing_species_matrix)) # arcsin squareroot transform
fringing_subset <- fringing[!fringing_rows_to_remove,] # make sure rows are also removed from dataframe
fringing_subset$stratum <- paste(fringing_subset$year,"-", fringing_subset$location, sep = "") # for merging with bray matrix later
brayfringing <- vegdist(fringing_species_matrix, "bray") # create bray dissimilarity matrix
# View(brayfringing)
fringing_braymat <- as.matrix(brayfringing) # conver to matrix object
# convert to pairwise comparison
fringing_comp <- fringing_braymat[lower.tri(fringing_braymat)] # extract lower triangle
fringing_row_indices <- row(fringing_braymat)[lower.tri(fringing_braymat)] # extract row indices for first quad in comparison
fringing_col_indices <- col(fringing_braymat)[lower.tri(fringing_braymat)] # extract col indices = second quad in comp
fringing_loc1 <- rownames(fringing_braymat)[fringing_row_indices] # populate appropriate vector
fringing_loc2 <- colnames(fringing_braymat)[fringing_col_indices] # populate appropriate vector
fringing_pairwise_df <- data.frame(loc1 = fringing_loc1, loc2 = fringing_loc2, dissimilarity = fringing_comp)
fringing_pairwise_df$year1 <- as.numeric(sub("(.*)-.*", "\\1", fringing_pairwise_df$loc1)) # extract timepoint from first quad in comp
fringing_pairwise_df$year2 <- as.numeric(sub("(.*)-.*", "\\1", fringing_pairwise_df$loc2)) # repeat for second quad
fringing_pairwise_df$time <- abs(fringing_pairwise_df$year2 - fringing_pairwise_df$year1) # extract length of differences
fringing_pairwise_df$location_1 <- sub(".*-(.*)", "\\1", fringing_pairwise_df$loc1) # extract quadrat id from first quad in comp
fringing_pairwise_df$location_2 <- sub(".*-(.*)", "\\1", fringing_pairwise_df$loc2) # repeat for second quad

@@

```{r backreef bray curtis}
backreef_species_matrix <- as.matrix(backreef[ , -which(names(backreef) %in% meta_cols)])
backreef_species_matrix <- backreef_species_matrix/100 # convert from percent to proportion
rownames(backreef_species_matrix) <- paste(backreef$year,"-", backreef$location, sep = "")
backreef_rows_to_remove <- rowSums(backreef_species_matrix) == 0 # remove rows with no algal species observed - thought this was done already
backreef_species_matrix <- backreef_species_matrix[!backreef_rows_to_remove,] 
backreef_species_matrix <- asin(sqrt(backreef_species_matrix)) # arcsin squareroot transform
backreef_subset <- backreef[!backreef_rows_to_remove,] # make sure rows are also removed from dataframe
backreef_subset$stratum <- paste(backreef_subset$year,"-", backreef_subset$location, sep = "") # for merging with bray matrix later
braybackreef <- vegdist(backreef_species_matrix, "bray") # create bray dissimilarity matrix
# View(braybackreef)
backreef_braymat <- as.matrix(braybackreef) # conver to matrix object
# convert to pairwise comparison
backreef_comp <- backreef_braymat[lower.tri(backreef_braymat)] # extract lower triangle
backreef_row_indices <- row(backreef_braymat)[lower.tri(backreef_braymat)] # extract row indices for first quad in comparison
backreef_col_indices <- col(backreef_braymat)[lower.tri(backreef_braymat)] # extract col indices = second quad in comp
backreef_loc1 <- rownames(backreef_braymat)[backreef_row_indices] # populate appropriate vector
backreef_loc2 <- colnames(backreef_braymat)[backreef_col_indices] # populate appropriate vector
backreef_pairwise_df <- data.frame(loc1 = backreef_loc1, loc2 = backreef_loc2, dissimilarity = backreef_comp)
backreef_pairwise_df$year1 <- as.numeric(sub("(.*)-.*", "\\1", backreef_pairwise_df$loc1)) # extract timepoint from first quad in comp
backreef_pairwise_df$year2 <- as.numeric(sub("(.*)-.*", "\\1", backreef_pairwise_df$loc2)) # repeat for second quad
backreef_pairwise_df$time <- abs(backreef_pairwise_df$year2 - backreef_pairwise_df$year1) # extract length of differences
backreef_pairwise_df$location_1 <- sub(".*-(.*)", "\\1", backreef_pairwise_df$loc1) # extract quadrat id from first quad in comp
backreef_pairwise_df$location_2 <- sub(".*-(.*)", "\\1", backreef_pairwise_df$loc2) # repeat for second quad

@@

```{r outer10 bray curtis}
outer10_species_matrix <- as.matrix(outer10[ , -which(names(outer10) %in% meta_cols)])
outer10_species_matrix <- outer10_species_matrix/100 # convert from percent to proportion
rownames(outer10_species_matrix) <- paste(outer10$year,"-", outer10$location, sep = "")
outer10_rows_to_remove <- rowSums(outer10_species_matrix) == 0 # remove rows with no algal species observed - thought this was done already
outer10_species_matrix <- outer10_species_matrix[!outer10_rows_to_remove,] 
outer10_species_matrix <- asin(sqrt(outer10_species_matrix)) # arcsin squareroot transform
outer10_subset <- outer10[!outer10_rows_to_remove,] # make sure rows are also removed from dataframe
outer10_subset$stratum <- paste(outer10_subset$year,"-", outer10_subset$location, sep = "") # for merging with bray matrix later
brayouter10 <- vegdist(outer10_species_matrix, "bray") # create bray dissimilarity matrix
# View(brayouter10)
outer10_braymat <- as.matrix(brayouter10) # conver to matrix object
# convert to pairwise comparison
outer10_comp <- outer10_braymat[lower.tri(outer10_braymat)] # extract lower triangle
outer10_row_indices <- row(outer10_braymat)[lower.tri(outer10_braymat)] # extract row indices for first quad in comparison
outer10_col_indices <- col(outer10_braymat)[lower.tri(outer10_braymat)] # extract col indices = second quad in comp
outer10_loc1 <- rownames(outer10_braymat)[outer10_row_indices] # populate appropriate vector
outer10_loc2 <- colnames(outer10_braymat)[outer10_col_indices] # populate appropriate vector
outer10_pairwise_df <- data.frame(loc1 = outer10_loc1, loc2 = outer10_loc2, dissimilarity = outer10_comp)
outer10_pairwise_df$year1 <- as.numeric(sub("(.*)-.*", "\\1", outer10_pairwise_df$loc1)) # extract timepoint from first quad in comp
outer10_pairwise_df$year2 <- as.numeric(sub("(.*)-.*", "\\1", outer10_pairwise_df$loc2)) # repeat for second quad
outer10_pairwise_df$time <- abs(outer10_pairwise_df$year2 - outer10_pairwise_df$year1) # extract length of differences
outer10_pairwise_df$location_1 <- sub(".*-(.*)", "\\1", outer10_pairwise_df$loc1) # extract quadrat id from first quad in comp
outer10_pairwise_df$location_2 <- sub(".*-(.*)", "\\1", outer10_pairwise_df$loc2) # repeat for second quad

@@

```{r outer17 bray curtis}
outer17_species_matrix <- as.matrix(outer17[ , -which(names(outer17) %in% meta_cols)])
outer17_species_matrix <- outer17_species_matrix/100 # convert from percent to proportion
rownames(outer17_species_matrix) <- paste(outer17$year,"-", outer17$location, sep = "")
outer17_rows_to_remove <- rowSums(outer17_species_matrix) == 0 # remove rows with no algal species observed - thought this was done already
outer17_species_matrix <- outer17_species_matrix[!outer17_rows_to_remove,] 
outer17_species_matrix <- asin(sqrt(outer17_species_matrix)) # arcsin squareroot transform
outer17_subset <- outer17[!outer17_rows_to_remove,] # make sure rows are also removed from dataframe
outer17_subset$stratum <- paste(outer17_subset$year,"-", outer17_subset$location, sep = "") # for merging with bray matrix later
brayouter17 <- vegdist(outer17_species_matrix, "bray") # create bray dissimilarity matrix
# View(brayouter17)
outer17_braymat <- as.matrix(brayouter17) # conver to matrix object
# convert to pairwise comparison
outer17_comp <- outer17_braymat[lower.tri(outer17_braymat)] # extract lower triangle
outer17_row_indices <- row(outer17_braymat)[lower.tri(outer17_braymat)] # extract row indices for first quad in comparison
outer17_col_indices <- col(outer17_braymat)[lower.tri(outer17_braymat)] # extract col indices = second quad in comp
outer17_loc1 <- rownames(outer17_braymat)[outer17_row_indices] # populate appropriate vector
outer17_loc2 <- colnames(outer17_braymat)[outer17_col_indices] # populate appropriate vector
outer17_pairwise_df <- data.frame(loc1 = outer17_loc1, loc2 = outer17_loc2, dissimilarity = outer17_comp)
outer17_pairwise_df$year1 <- as.numeric(sub("(.*)-.*", "\\1", outer17_pairwise_df$loc1)) # extract timepoint from first quad in comp
outer17_pairwise_df$year2 <- as.numeric(sub("(.*)-.*", "\\1", outer17_pairwise_df$loc2)) # repeat for second quad
outer17_pairwise_df$time <- abs(outer17_pairwise_df$year2 - outer17_pairwise_df$year1) # extract length of differences
outer17_pairwise_df$location_1 <- sub(".*-(.*)", "\\1", outer17_pairwise_df$loc1) # extract quadrat id from first quad in comp
outer17_pairwise_df$location_2 <- sub(".*-(.*)", "\\1", outer17_pairwise_df$loc2) # repeat for second quad

@@

@@

# Smithson verkuilen transformation for beta-regression
sv_trans = function(prop, N, s = 0.000005){
  (prop*(nrow(N) - 1) + s)/nrow(N)
  # where prop is the proportional value you're transforming, 
  # N is the sample size, which is specified by taking the number of rows/observations from a given dataframe,
  # and s is a small offset 
}

# transformed dissimilarities
bray_time_series$dissimilarity_trans <- sv_trans(bray_time_series$dissimilarity, bray_time_series)

@@

emtrends(bray_mod_universal, pairwise ~ habitat |island_side, "time")
bray_trends_emm <- emmip(bray_mod_universal, habitat ~ time | island_side, at = list(time = seq(1,16,1)), plotit = F, CIs = T)
inv_logit <- function(x) {
  exp(x)/(1+exp(x))
}
```

@@

@@

#### WARNING: THIS CHUNK TAKES ~2 DAYS TO RUN! SUGGESTEDF TO USE PROVIDED CSV FILES WITH ORDINATION DATA ######

>>>>

not_algae <- c("ascidian", "bare_space", "coral", "coral_rubble", "corallimorpharia", "heteractis_sp", "millepora_platyphylla", "no_data", "sand", "sarcophyton_sp", "shell_debris", "soft_coral", "sponge", "tridacna_sp", "zooxanthellae", "row_sums")

@@

```{r fringing bray curtis}
fringing_species_matrix <- as.matrix(fringing[ , -which(names(fringing) %in% meta_cols)])
fringing_species_matrix <- fringing_species_matrix/100 # convert from percent to proportion
rownames(fringing_species_matrix) <- paste(fringing$year,"-", fringing$location, sep = "")

# filter rare species
fringing_species_presence <- colSums(fringing_species_matrix > 0) # Presence absence matrix
fringing_species_percentage <- (fringing_species_presence/nrow(fringing_species_matrix)) # proportion of rows where a species is present
fringing_common_species <- names(fringing_species_percentage[fringing_species_percentage > 0.05]) # filter for presence in more than 5% of quads - only 9 species...
fringing_species_matrix <- fringing_species_matrix[,colnames(fringing_species_matrix) %in% fringing_common_species] # filter original species matrix

fringing_rows_to_remove <- rowSums(fringing_species_matrix) == 0 # remove rows with no algal species observed - thought this was done already
fringing_species_matrix <- fringing_species_matrix[!fringing_rows_to_remove,]

fringing_species_matrix_untransformed <- fringing_species_matrix
fringing_species_matrix_transformed <- asin(sqrt(fringing_species_matrix)) # arcsin squareroot transform
fringing_subset <- fringing[!fringing_rows_to_remove,] # make sure rows are also removed from dataframe
fringing_subset$stratum <- paste(fringing_subset$year,"-", fringing_subset$location, sep = "") # for merging with bray matrix later

brayfringing <- vegdist(fringing_species_matrix_transformed, "bray") # create bray dissimilarity matrix
# View(brayfringing)
brayfringing_untransformed <- vegdist(fringing_species_matrix_untransformed, "bray")
fringing_braymat <- as.matrix(brayfringing) # conver to matrix object
# convert to pairwise comparison
fringing_comp <- fringing_braymat[lower.tri(fringing_braymat)] # extract lower triangle
fringing_row_indices <- row(fringing_braymat)[lower.tri(fringing_braymat)] # extract row indices for first quad in comparison
fringing_col_indices <- col(fringing_braymat)[lower.tri(fringing_braymat)] # extract col indices = second quad in comp
fringing_loc1 <- rownames(fringing_braymat)[fringing_row_indices] # populate appropriate vector
fringing_loc2 <- colnames(fringing_braymat)[fringing_col_indices] # populate appropriate vector
fringing_pairwise_df <- data.frame(loc1 = fringing_loc1, loc2 = fringing_loc2, dissimilarity = fringing_comp)
fringing_pairwise_df$year1 <- as.numeric(sub("(.*)-.*", "\\1", fringing_pairwise_df$loc1)) # extract timepoint from first quad in comp
fringing_pairwise_df$year2 <- as.numeric(sub("(.*)-.*", "\\1", fringing_pairwise_df$loc2)) # repeat for second quad
fringing_pairwise_df$time <- abs(fringing_pairwise_df$year2 - fringing_pairwise_df$year1) # extract length of differences
fringing_pairwise_df$location_1 <- sub(".*-(.*)", "\\1", fringing_pairwise_df$loc1) # extract quadrat id from first quad in comp
fringing_pairwise_df$location_2 <- sub(".*-(.*)", "\\1", fringing_pairwise_df$loc2) # repeat for second quad

fringing_untransformed_braymat <- as.matrix(brayfringing_untransformed) # conver to matrix object
# convert to pairwise comparison
fringing_untransformed_comp <- fringing_untransformed_braymat[lower.tri(fringing_untransformed_braymat)] # extract lower triangle
fringing_untransformed_row_indices <- row(fringing_untransformed_braymat)[lower.tri(fringing_untransformed_braymat)] # extract row indices for first quad in comparison
fringing_untransformed_col_indices <- col(fringing_untransformed_braymat)[lower.tri(fringing_untransformed_braymat)] # extract col indices = second quad in comp
fringing_untransformed_loc1 <- rownames(fringing_untransformed_braymat)[fringing_untransformed_row_indices] # populate appropriate vector
fringing_untransformed_loc2 <- colnames(fringing_untransformed_braymat)[fringing_untransformed_col_indices] # populate appropriate vector
fringing_untransformed_pairwise_df <- data.frame(loc1 = fringing_untransformed_loc1, loc2 = fringing_untransformed_loc2, dissimilarity = fringing_untransformed_comp)
fringing_untransformed_pairwise_df$year1 <- as.numeric(sub("(.*)-.*", "\\1", fringing_untransformed_pairwise_df$loc1)) # extract timepoint from first quad in comp
fringing_untransformed_pairwise_df$year2 <- as.numeric(sub("(.*)-.*", "\\1", fringing_untransformed_pairwise_df$loc2)) # repeat for second quad
fringing_untransformed_pairwise_df$time <- abs(fringing_untransformed_pairwise_df$year2 - fringing_untransformed_pairwise_df$year1) # extract length of differences
fringing_untransformed_pairwise_df$location_1 <- sub(".*-(.*)", "\\1", fringing_untransformed_pairwise_df$loc1) # extract quadrat id from first quad in comp
fringing_untransformed_pairwise_df$location_2 <- sub(".*-(.*)", "\\1", fringing_untransformed_pairwise_df$loc2) # repeat for second quad

@@

```{r backreef bray curtis}
backreef_species_matrix <- as.matrix(backreef[ , -which(names(backreef) %in% meta_cols)])
backreef_species_matrix <- backreef_species_matrix/100 # convert from percent to proportion
rownames(backreef_species_matrix) <- paste(backreef$year,"-", backreef$location, sep = "")

# filter rare species
backreef_species_presence <- colSums(backreef_species_matrix > 0) # Presence absence matrix
backreef_species_percentage <- (backreef_species_presence/nrow(backreef_species_matrix)) # proportion of rows where a species is present
backreef_common_species <- names(backreef_species_percentage[backreef_species_percentage > 0.05]) # filter for presence in more than 5% of quads - only 9 species...
backreef_species_matrix <- backreef_species_matrix[,colnames(backreef_species_matrix) %in% backreef_common_species] # filter original species matrix

backreef_rows_to_remove <- rowSums(backreef_species_matrix) == 0 # remove rows with no algal species observed - thought this was done already
backreef_species_matrix <- backreef_species_matrix[!backreef_rows_to_remove,]

backreef_species_matrix_untransformed <- backreef_species_matrix
backreef_species_matrix_transformed <- asin(sqrt(backreef_species_matrix)) # arcsin squareroot transform
backreef_subset <- backreef[!backreef_rows_to_remove,] # make sure rows are also removed from dataframe
backreef_subset$stratum <- paste(backreef_subset$year,"-", backreef_subset$location, sep = "") # for merging with bray matrix later
braybackreef <- vegdist(backreef_species_matrix_transformed, "bray") # create bray dissimilarity matrix
# View(braybackreef)
braybackreef_untransformed <- vegdist(backreef_species_matrix_untransformed, "bray")
backreef_braymat <- as.matrix(braybackreef) # conver to matrix object
# convert to pairwise comparison
backreef_comp <- backreef_braymat[lower.tri(backreef_braymat)] # extract lower triangle
backreef_row_indices <- row(backreef_braymat)[lower.tri(backreef_braymat)] # extract row indices for first quad in comparison
backreef_col_indices <- col(backreef_braymat)[lower.tri(backreef_braymat)] # extract col indices = second quad in comp
backreef_loc1 <- rownames(backreef_braymat)[backreef_row_indices] # populate appropriate vector
backreef_loc2 <- colnames(backreef_braymat)[backreef_col_indices] # populate appropriate vector
backreef_pairwise_df <- data.frame(loc1 = backreef_loc1, loc2 = backreef_loc2, dissimilarity = backreef_comp)
backreef_pairwise_df$year1 <- as.numeric(sub("(.*)-.*", "\\1", backreef_pairwise_df$loc1)) # extract timepoint from first quad in comp
backreef_pairwise_df$year2 <- as.numeric(sub("(.*)-.*", "\\1", backreef_pairwise_df$loc2)) # repeat for second quad
backreef_pairwise_df$time <- abs(backreef_pairwise_df$year2 - backreef_pairwise_df$year1) # extract length of differences
backreef_pairwise_df$location_1 <- sub(".*-(.*)", "\\1", backreef_pairwise_df$loc1) # extract quadrat id from first quad in comp
backreef_pairwise_df$location_2 <- sub(".*-(.*)", "\\1", backreef_pairwise_df$loc2) # repeat for second quad

backreef_untransformed_braymat <- as.matrix(braybackreef_untransformed) # conver to matrix object
# convert to pairwise comparison
backreef_untransformed_comp <- backreef_untransformed_braymat[lower.tri(backreef_untransformed_braymat)] # extract lower triangle
backreef_untransformed_row_indices <- row(backreef_untransformed_braymat)[lower.tri(backreef_untransformed_braymat)] # extract row indices for first quad in comparison
backreef_untransformed_col_indices <- col(backreef_untransformed_braymat)[lower.tri(backreef_untransformed_braymat)] # extract col indices = second quad in comp
backreef_untransformed_loc1 <- rownames(backreef_untransformed_braymat)[backreef_untransformed_row_indices] # populate appropriate vector
backreef_untransformed_loc2 <- colnames(backreef_untransformed_braymat)[backreef_untransformed_col_indices] # populate appropriate vector
backreef_untransformed_pairwise_df <- data.frame(loc1 = backreef_untransformed_loc1, loc2 = backreef_untransformed_loc2, dissimilarity = backreef_untransformed_comp)
backreef_untransformed_pairwise_df$year1 <- as.numeric(sub("(.*)-.*", "\\1", backreef_untransformed_pairwise_df$loc1)) # extract timepoint from first quad in comp
backreef_untransformed_pairwise_df$year2 <- as.numeric(sub("(.*)-.*", "\\1", backreef_untransformed_pairwise_df$loc2)) # repeat for second quad
backreef_untransformed_pairwise_df$time <- abs(backreef_untransformed_pairwise_df$year2 - backreef_untransformed_pairwise_df$year1) # extract length of differences
backreef_untransformed_pairwise_df$location_1 <- sub(".*-(.*)", "\\1", backreef_untransformed_pairwise_df$loc1) # extract quadrat id from first quad in comp
backreef_untransformed_pairwise_df$location_2 <- sub(".*-(.*)", "\\1", backreef_untransformed_pairwise_df$loc2) # repeat for second quad

@@

```{r outer10 bray curtis}
outer10_species_matrix <- as.matrix(outer10[ , -which(names(outer10) %in% meta_cols)])
outer10_species_matrix <- outer10_species_matrix/100 # convert from percent to proportion
rownames(outer10_species_matrix) <- paste(outer10$year,"-", outer10$location, sep = "")

# filter rare species
outer10_species_presence <- colSums(outer10_species_matrix > 0) # Presence absence matrix
outer10_species_percentage <- (outer10_species_presence/nrow(outer10_species_matrix)) # proportion of rows where a species is present
outer10_common_species <- names(outer10_species_percentage[outer10_species_percentage > 0.05]) # filter for presence in more than 5% of quads - only 9 species...
outer10_species_matrix <- outer10_species_matrix[,colnames(outer10_species_matrix) %in% outer10_common_species] # filter original species matrix

outer10_rows_to_remove <- rowSums(outer10_species_matrix) == 0 # remove rows with no algal species observed - thought this was done already
outer10_species_matrix <- outer10_species_matrix[!outer10_rows_to_remove,]

outer10_species_matrix_untransformed <- outer10_species_matrix
outer10_species_matrix_transformed <- asin(sqrt(outer10_species_matrix)) # arcsin squareroot transform
outer10_subset <- outer10[!outer10_rows_to_remove,] # make sure rows are also removed from dataframe
outer10_subset$stratum <- paste(outer10_subset$year,"-", outer10_subset$location, sep = "") # for merging with bray matrix later
brayouter10 <- vegdist(outer10_species_matrix_transformed, "bray") # create bray dissimilarity matrix
# View(brayouter10)
brayouter10_untransformed <- vegdist(outer10_species_matrix_untransformed, "bray")
outer10_braymat <- as.matrix(brayouter10) # conver to matrix object
# convert to pairwise comparison
outer10_comp <- outer10_braymat[lower.tri(outer10_braymat)] # extract lower triangle
outer10_row_indices <- row(outer10_braymat)[lower.tri(outer10_braymat)] # extract row indices for first quad in comparison
outer10_col_indices <- col(outer10_braymat)[lower.tri(outer10_braymat)] # extract col indices = second quad in comp
outer10_loc1 <- rownames(outer10_braymat)[outer10_row_indices] # populate appropriate vector
outer10_loc2 <- colnames(outer10_braymat)[outer10_col_indices] # populate appropriate vector
outer10_pairwise_df <- data.frame(loc1 = outer10_loc1, loc2 = outer10_loc2, dissimilarity = outer10_comp)
outer10_pairwise_df$year1 <- as.numeric(sub("(.*)-.*", "\\1", outer10_pairwise_df$loc1)) # extract timepoint from first quad in comp
outer10_pairwise_df$year2 <- as.numeric(sub("(.*)-.*", "\\1", outer10_pairwise_df$loc2)) # repeat for second quad
outer10_pairwise_df$time <- abs(outer10_pairwise_df$year2 - outer10_pairwise_df$year1) # extract length of differences
outer10_pairwise_df$location_1 <- sub(".*-(.*)", "\\1", outer10_pairwise_df$loc1) # extract quadrat id from first quad in comp
outer10_pairwise_df$location_2 <- sub(".*-(.*)", "\\1", outer10_pairwise_df$loc2) # repeat for second quad

outer10_untransformed_braymat <- as.matrix(brayouter10_untransformed) # conver to matrix object
# convert to pairwise comparison
outer10_untransformed_comp <- outer10_untransformed_braymat[lower.tri(outer10_untransformed_braymat)] # extract lower triangle
outer10_untransformed_row_indices <- row(outer10_untransformed_braymat)[lower.tri(outer10_untransformed_braymat)] # extract row indices for first quad in comparison
outer10_untransformed_col_indices <- col(outer10_untransformed_braymat)[lower.tri(outer10_untransformed_braymat)] # extract col indices = second quad in comp
outer10_untransformed_loc1 <- rownames(outer10_untransformed_braymat)[outer10_untransformed_row_indices] # populate appropriate vector
outer10_untransformed_loc2 <- colnames(outer10_untransformed_braymat)[outer10_untransformed_col_indices] # populate appropriate vector
outer10_untransformed_pairwise_df <- data.frame(loc1 = outer10_untransformed_loc1, loc2 = outer10_untransformed_loc2, dissimilarity = outer10_untransformed_comp)
outer10_untransformed_pairwise_df$year1 <- as.numeric(sub("(.*)-.*", "\\1", outer10_untransformed_pairwise_df$loc1)) # extract timepoint from first quad in comp
outer10_untransformed_pairwise_df$year2 <- as.numeric(sub("(.*)-.*", "\\1", outer10_untransformed_pairwise_df$loc2)) # repeat for second quad
outer10_untransformed_pairwise_df$time <- abs(outer10_untransformed_pairwise_df$year2 - outer10_untransformed_pairwise_df$year1) # extract length of differences
outer10_untransformed_pairwise_df$location_1 <- sub(".*-(.*)", "\\1", outer10_untransformed_pairwise_df$loc1) # extract quadrat id from first quad in comp
outer10_untransformed_pairwise_df$location_2 <- sub(".*-(.*)", "\\1", outer10_untransformed_pairwise_df$loc2) # repeat for second quad

@@

```{r outer17 bray curtis}
outer17_species_matrix <- as.matrix(outer17[ , -which(names(outer17) %in% meta_cols)])
outer17_species_matrix <- outer17_species_matrix/100 # convert from percent to proportion
rownames(outer17_species_matrix) <- paste(outer17$year,"-", outer17$location, sep = "")

# filter rare species
outer17_species_presence <- colSums(outer17_species_matrix > 0) # Presence absence matrix
outer17_species_percentage <- (outer17_species_presence/nrow(outer17_species_matrix)) # proportion of rows where a species is present
outer17_common_species <- names(outer17_species_percentage[outer17_species_percentage > 0.05]) # filter for presence in more than 5% of quads - only 9 species...
outer17_species_matrix <- outer17_species_matrix[,colnames(outer17_species_matrix) %in% outer17_common_species] # filter original species matrix

outer17_rows_to_remove <- rowSums(outer17_species_matrix) == 0 # remove rows with no algal species observed - thought this was done already
outer17_species_matrix <- outer17_species_matrix[!outer17_rows_to_remove,]

outer17_species_matrix_untransformed <- outer17_species_matrix
outer17_species_matrix_transformed <- asin(sqrt(outer17_species_matrix)) # arcsin squareroot transform
outer17_subset <- outer17[!outer17_rows_to_remove,] # make sure rows are also removed from dataframe
outer17_subset$stratum <- paste(outer17_subset$year,"-", outer17_subset$location, sep = "") # for merging with bray matrix later
brayouter17 <- vegdist(outer17_species_matrix_transformed, "bray") # create bray dissimilarity matrix
# View(brayouter17)
brayouter17_untransformed <- vegdist(outer17_species_matrix_untransformed, "bray")
outer17_braymat <- as.matrix(brayouter17) # conver to matrix object
# convert to pairwise comparison
outer17_comp <- outer17_braymat[lower.tri(outer17_braymat)] # extract lower triangle
outer17_row_indices <- row(outer17_braymat)[lower.tri(outer17_braymat)] # extract row indices for first quad in comparison
outer17_col_indices <- col(outer17_braymat)[lower.tri(outer17_braymat)] # extract col indices = second quad in comp
outer17_loc1 <- rownames(outer17_braymat)[outer17_row_indices] # populate appropriate vector
outer17_loc2 <- colnames(outer17_braymat)[outer17_col_indices] # populate appropriate vector
outer17_pairwise_df <- data.frame(loc1 = outer17_loc1, loc2 = outer17_loc2, dissimilarity = outer17_comp)
outer17_pairwise_df$year1 <- as.numeric(sub("(.*)-.*", "\\1", outer17_pairwise_df$loc1)) # extract timepoint from first quad in comp
outer17_pairwise_df$year2 <- as.numeric(sub("(.*)-.*", "\\1", outer17_pairwise_df$loc2)) # repeat for second quad
outer17_pairwise_df$time <- abs(outer17_pairwise_df$year2 - outer17_pairwise_df$year1) # extract length of differences
outer17_pairwise_df$location_1 <- sub(".*-(.*)", "\\1", outer17_pairwise_df$loc1) # extract quadrat id from first quad in comp
outer17_pairwise_df$location_2 <- sub(".*-(.*)", "\\1", outer17_pairwise_df$loc2) # repeat for second quad

outer17_untransformed_braymat <- as.matrix(brayouter17_untransformed) # conver to matrix object
# convert to pairwise comparison
outer17_untransformed_comp <- outer17_untransformed_braymat[lower.tri(outer17_untransformed_braymat)] # extract lower triangle
outer17_untransformed_row_indices <- row(outer17_untransformed_braymat)[lower.tri(outer17_untransformed_braymat)] # extract row indices for first quad in comparison
outer17_untransformed_col_indices <- col(outer17_untransformed_braymat)[lower.tri(outer17_untransformed_braymat)] # extract col indices = second quad in comp
outer17_untransformed_loc1 <- rownames(outer17_untransformed_braymat)[outer17_untransformed_row_indices] # populate appropriate vector
outer17_untransformed_loc2 <- colnames(outer17_untransformed_braymat)[outer17_untransformed_col_indices] # populate appropriate vector
outer17_untransformed_pairwise_df <- data.frame(loc1 = outer17_untransformed_loc1, loc2 = outer17_untransformed_loc2, dissimilarity = outer17_untransformed_comp)
outer17_untransformed_pairwise_df$year1 <- as.numeric(sub("(.*)-.*", "\\1", outer17_untransformed_pairwise_df$loc1)) # extract timepoint from first quad in comp
outer17_untransformed_pairwise_df$year2 <- as.numeric(sub("(.*)-.*", "\\1", outer17_untransformed_pairwise_df$loc2)) # repeat for second quad
outer17_untransformed_pairwise_df$time <- abs(outer17_untransformed_pairwise_df$year2 - outer17_untransformed_pairwise_df$year1) # extract length of differences
outer17_untransformed_pairwise_df$location_1 <- sub(".*-(.*)", "\\1", outer17_untransformed_pairwise_df$loc1) # extract quadrat id from first quad in comp
outer17_untransformed_pairwise_df$location_2 <- sub(".*-(.*)", "\\1", outer17_untransformed_pairwise_df$loc2) # repeat for second quad

@@

fringing_untransformed_pairwise_subset <- fringing_untransformed_pairwise_df[fringing_untransformed_pairwise_df$location_1 == fringing_untransformed_pairwise_df$location_2,]
backreef_untransformed_pairwise_subset <- backreef_untransformed_pairwise_df[backreef_untransformed_pairwise_df$location_1 == backreef_untransformed_pairwise_df$location_2,]
outer10_untransformed_pairwise_subset <- outer10_untransformed_pairwise_df[outer10_untransformed_pairwise_df$location_1 == outer10_untransformed_pairwise_df$location_2,]
outer17_untransformed_pairwise_subset <- outer17_untransformed_pairwise_df[outer17_untransformed_pairwise_df$location_1 == outer17_untransformed_pairwise_df$location_2,]

comb_untransformed_subset <- rbind(fringing_untransformed_pairwise_subset,
                     backreef_untransformed_pairwise_subset,
                     outer10_untransformed_pairwise_subset,
                     outer17_untransformed_pairwise_subset)
# location_1 and location_2 columns are all the same now
# create new col just called "location" for merge with other metadata
comb_untransformed_subset$location <- comb_untransformed_subset$location_1
bray_untransformed_time_series <- merge(meta_df, comb_untransformed_subset, by = "location")

@@

# Smithson verkuilen transformation for beta-regression
sv_trans <- function(prop, N, s = 0.000005){
  (prop*(nrow(N) - 1) + s)/nrow(N)
  # where prop is the proportional value you're transforming, 
  # N is the sample size, which is specified by taking the number of rows/observations from a given dataframe,
  # and s is a small offset 
}

inv_logit <- function(x) {
  exp(x)/(1+exp(x))
}

# transformed dissimilarities
bray_time_series$dissimilarity_trans <- sv_trans(bray_time_series$dissimilarity, bray_time_series)
bray_untransformed_time_series$dissimilarity_trans <- sv_trans(bray_untransformed_time_series$dissimilarity, bray_time_series)

@@

emtrends(bray_mod_universal, pairwise ~ habitat |island_side, "time")
bray_trends_emm <- emmip(bray_mod_universal, habitat ~ time | island_side, at = list(time = seq(1,16,1)), plotit = F, CIs = T)

bray_untransformed_mod_universal <- glmmTMB(dissimilarity_trans ~ time*habitat*island_side + (1 | island_side/site/transect/location),
                              na.action = na.pass, family = beta_family(link = "logit"), data = bray_untransformed_time_series) #
# model.sel(dredge(bray_untransformed_mod_universal)) # top model as specified
summary(bray_untransformed_mod_universal) # a fair bit of variance w/i quads/transects
hist(residuals(bray_untransformed_mod_universal))
plot(predict(bray_untransformed_mod_universal) ~ residuals(bray_untransformed_mod_universal)) # looks like reflection of logit curve

emtrends(bray_untransformed_mod_universal, pairwise ~ habitat |island_side, "time")
bray_untransformed_trends_emm <- emmip(bray_untransformed_mod_universal, habitat ~ time | island_side, at = list(time = seq(1,16,1)), plotit = F, CIs = T)

@@

ggplot()+
   geom_point(data = bray_untransformed_time_series, aes(x = time, y = dissimilarity, colour = island_side), alpha = 0.01 ) +
   geom_line(data = bray_untransformed_trends_emm, aes(x = time, y = inv_logit(yvar), colour = island_side), linewidth = 2) +
   geom_ribbon(data = bray_untransformed_trends_emm, aes(x = time, ymin = inv_logit(LCL), ymax = inv_logit(UCL), fill = island_side), alpha = 0.3) +
   facet_wrap(~habitat) +

@@

#### WARNING: THIS CHUNK TAKES ~2 DAYS TO RUN! SUGGESTED TO USE PROVIDED CSV FILES WITH ORDINATION DATA ######
